Synthesizing Quotient Lenses

### 2018-03-15 papers/QuotientLenses/main.tex
Quick future work paragraphs

\section{Introduction}
Programmers often need to write programs that bidirectionally convert data
between two different formats. A time-tested and appealing way to do this is
to use \emph{lenses}~\cite{Lenses}, programs that simultaneously define pairs of
functions for translating the data from the source format to the target format
(the \emph{get} direction) and back again (the \emph{put} direction). The
main reason that lenses are appealing is because lens programming languages
guarantee that the \kw{put} and \kw{get} functions of a lens satisfy
certain \emph{lens laws} constraining how the two functions interact with each
other.

@@

For example, consider the following QRE for writing author
names:\bcp{Solomon: Someone noted on their copy of the intro from
  class yesterday that this example is backwards.  Did you go over the
  marked-up copies that I gave you?}\sam{I'm not sure what this comment means
  and also I forgot those notes in the office. I'll go through them more
  thoroughly tomorrow morning.}
%
\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

The denotation $\llbracket q \rrbracket$ of a QRE lens $q : Q_1 \Leftrightarrow
Q_2$ is a quotient lens $\llbracket q \rrbracket : W(Q_1)/{\eqrel{Q_1}}
\Longleftrightarrow W(Q_2)/{\eqrel{Q_2}}$. The typing rules and denotation of
QRE lenses are given in Figure~\ref{fig:qlenssemantics}.
\begin{figure}[ht]
\centering
\[
\begin{array}[b]{l@{\qquad}l}
\RuleSide{\ell : R \Leftrightarrow S}{\kw{lift}(\ell): id(R) \Leftrightarrow
id(S)}{\mathit{(Lift)}} &
\begin{array}[b]{rcl}
\llbracket \kw{lift}(\ell) \rrbracket.\get &=&  \llbracket \ell \rrbracket\\
\llbracket \kw{lift}(\ell) \rrbracket.\lput &=& \llbracket \ell \rrbracket^{-1}
\end{array}
\end{array}
\]
\[
\begin{array}[b]{c@{\qquad}c}
\RuleSide{q : Q_2  \Leftrightarrow Q_3 &
\wf{Q_1} &
K(Q_1) = W(Q_2)}
{\lquot(Q_1, q): Q_1 \; ; \; Q_2 \Leftrightarrow Q_3}{\mathit(Lquot)} &
\begin{array}[b]{rcl}
\llbracket \lquot(Q_1, q) \rrbracket.\get  &=& \llbracket q
\rrbracket.\get \circ \canonize(Q_1)\\
\llbracket \lquot(Q_1, q) \rrbracket.\lput &=& \llbracket q
\rrbracket.\lput
\end{array}
\end{array}
\]

\[
\begin{array}[b]{c@{\qquad}c}
\RuleSide{q : Q_1 \Leftrightarrow Q_3 & \wf{Q_3} & W(Q_3) = K(Q_2)}
{\rquot(q, Q_1):Q_1 \Leftrightarrow Q_2 \; ; \; Q_3}{\mathit{(Rquot)}} &
\begin{array}[b]{rcl}
\llbracket \rquot(q, Q_1) \rrbracket.\get  &=& \llbracket q
\rrbracket.\get\\
\llbracket \rquot(q, Q_1) \rrbracket.\lput &=& \llbracket q
\rrbracket.\lput \circ \canonize(Q_3)
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{q : Q_1 \Leftrightarrow Q_2 &
W(Q_1)^{*!},W(Q_2)^{*!} & K(Q_1)^{*!}, K(Q_2)^{*!}}
{{q}^* : {Q_1}^* \Leftrightarrow {Q_2}^*}{\mathit{(Star)}} &
\begin{array}[b]{rcl}
\llbracket {q}^* \rrbracket.\get  &=& (\llbracket q \rrbracket.\get)^*\\
\llbracket {q}^* \rrbracket.\lput &=& (\llbracket q \rrbracket.\get)^*
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{\substack{q_1 : Q_1 \Leftrightarrow Q_3 \\ q_2 : Q_2 \Leftrightarrow
Q_4} & {\substack{W(Q_1) \cdot^! W(Q_2)\\ K(Q_1) \cdot^! K(Q_2)}}
& {\substack{W(Q_3) \cdot^! W(Q_4)\\ K(Q_3) \cdot^! K(Q_4)}}}
{q_1 \cdot q_2: Q_1 \cdot Q_2 \Leftrightarrow Q_3 \cdot Q_4}{\mathit{(Concat)}}
&
\begin{array}[b]{rcl}
\llbracket q_1 \cdot q_2 \rrbracket.\get  &=& \llbracket q_1 \rrbracket.\get
\cdot \llbracket q_2 \rrbracket.\get\\
\llbracket q_1 \cdot q_2\rrbracket.\lput &=& \llbracket q_1 \rrbracket.\lput
\cdot \llbracket q_2 \rrbracket.\lput
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{\substack{q_1 : Q_1 \Leftrightarrow Q_3 \\ q_2 : Q_2 \Leftrightarrow
Q_4} & {\substack{W(Q_1) \cdot^! W(Q_2)\\ K(Q_1) \cdot^! K(Q_2)}}
& {\substack{W(Q_4) \cdot^! W(Q_3)\\ K(Q_4) \cdot^! K(Q_3)}}}
{q= q_1 \cdot q_2: Q_1 \cdot Q_2 \Leftrightarrow Q_4 \cdot
Q_3}{\mathit{(Swap)}} &
\begin{array}[b]{rcl}
\llbracket q \rrbracket.\get(s_1 \cdot s_2)  &=& \llbracket q_2
\rrbracket.\get(s_2) \cdot \llbracket q_1 \rrbracket.\get(s_1)\\
\llbracket q \rrbracket.\lput(t_2 \cdot t_1) &=&
\llbracket q_1 \rrbracket.\lput(t_1) \cdot \llbracket q_2 \rrbracket.\lput(t_2)
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{\substack{q_1 : Q_1 \Leftrightarrow Q_3 \\ q_2 : Q_2 \Leftrightarrow
Q_4} & \substack{\mathcal{L}(W(Q_1)) \cap \mathcal{L}(W(Q_2)) = \varnothing
\\ \mathcal{L}(W(Q_3)) \cap \mathcal{L}(W(Q_4)) = \varnothing }}
{q_1 \sep q_2: (Q_1 \sep Q_2)\Leftrightarrow (Q_3 \sep Q_4)}{\mathit{(Or)}} &
\begin{array}[b]{rcl}
\llbracket q_1 \sep q_2 \rrbracket.\get(s) =
\begin{cases}
\llbracket q_1 \rrbracket.\get (s) & \text{if } s \in \mathcal{L}(W(Q_1))\\
\llbracket q_2 \rrbracket.\get (s) & \text{if } s \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\llbracket q_1 \sep q_2 \rrbracket.\lput(s) =
\begin{cases}
\llbracket q_1 \rrbracket.\lput (s) & \text{if } s \in \mathcal{L}(W(Q_3))\\
\llbracket q_2 \rrbracket.\lput (s) & \text{if } s \in \mathcal{L}(W(Q_4))\\
\end{cases}
\end{array}
\end{array}
\]

@@

By the induction hypothesis, there exist bijective lenses,
$\ell_1 : K(Q_1) \Leftrightarrow K(Q_2)$ and $\ell_2 : K(Q_3) \Leftrightarrow
K(Q_4)$ such that
\begin{center}:
\begin{tabular}{l l}
$\llbracket q_1 \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(Q_1)$ & $\llbracket q_2 \rrbracket.\get = \llbracket \ell_2
\rrbracket \circ \canonize(Q_3)$ \\
$\llbracket q_1 \rrbracket.\lput = {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_2)$ & $\llbracket q_2 \rrbracket.\lput = {\llbracket \ell_2
\rrbracket}^{-1} \circ \canonize(Q_4)$
\end{tabular}
\end{center}
%
Consequently:
\begin{align*}
\llbracket q_2 \rrbracket.\get \circ \llbracket q_1 \rrbracket.\get &=
(\llbracket \ell_2 \rrbracket \circ \canonize(Q_3)) \circ (\llbracket \ell_1
\rrbracket \circ \canonize(Q_1))\\
&= \llbracket \ell_2 \rrbracket \circ (\canonize(Q_3) \circ \llbracket \ell_1
\rrbracket) \circ \canonize(Q_1)\\
&= (\llbracket \ell_2 \rrbracket \circ \llbracket \ell_1 \rrbracket) \circ
\canonize(Q_1)\\
&= \llbracket \ell_1 \; ; \; \ell_2 \rrbracket \circ \canonize(Q_1)
\end{align*}
We are permitted to claim the third step from the second since
$\canonize(Q_3)$ is the identity function on $K(Q_2)$ which is
syntactically equal to $K(Q_3)$ by assumption. A similar argument shows that:
$$\llbracket q_1 \rrbracket.\lput \circ \llbracket q_2 \rrbracket.\lput =
\llbracket \ell_1 \; ; \; \ell_2 \rrbracket^{-1} \circ
\canonize(Q_4)$$

@@

\section{Conclusion and Future Work}
In this paper, we showed how to synthesize a class of bijective quotient lenses
using the bijective lens synthesis system Optician as a plug-in for our
synthesis algorithm. In order to achieve this, we first introduced {\em
Quotient regular languages} to specify regular languages with an equivalence
relation defined on them. Then, we introduced {\em QRE lenses}, which
are bijective quotient lenses that map between QREs via bijective lenses. We
proved a normal form theorem for QREs that enabled us to (1) extend the
synthesis algorithm used by Optician to synthesize QRE lenses, and (2) prove
that if there is a QRE lens that satisfies the input specification, then the
extended algorithm returns such a lens. Finally we tested QRE-enhanced optician
on the Optician benchmark suite and demonstrated that QREs can save programmers
significant effort compared to both Optician and basic Boomerang, and that
QRE-enhanced Optician maintains competitive performance over both of these
systems.

\bcp{And future work!}  \saz{I'm lukewarm about including a description of
future work.  Unless there are specific conjectures or open problems that
are worth mentioning, I'm not sure that it's really that important.}
\bcp{Agreed. But don't we have any ideas for how to extend or generalize or
things to do next or additional questions to consider??  E.g.: 
\begin{itemize}
\item is it easy to extend all this to
asymmetric and symmetric lenses?  (If so, we should say so!)
\item What properties do the bijective lens combinators have to have in
order to validate the normalization property?  (I.e., if we tried to extend
Boomerang's bijective lens combinators with another one, what would happen?)
\item Do we have more ideas for QRE primitives?
\item Etc.
\end{itemize}
}
\label{concl}

>>>>

\section{Introduction}
Programmers often need to write programs that bidirectionally convert data
between two different formats.\bcp{It's a shame to lose the concrete
  examples here.} A time-tested and appealing way to do this is
to use \emph{lenses}~\cite{Lenses}, programs that simultaneously define pairs of
functions for translating the data from the source format to the target format
(the \emph{get} direction) and back again (the \emph{put} direction). The
main reason that lenses are appealing is because lens programming languages
guarantee that the \kw{put} and \kw{get} functions of a lens satisfy
certain \emph{lens laws} constraining how the two functions interact with each
other.

@@

For example, consider the following QRE for writing author
names:\bcp{Solomon: Someone noted on their copy of the intro from
  class yesterday that this example is backwards.  Did you go over the
  marked-up copies that I gave you?}\sam{I'm not sure what this comment means
  and also I forgot those notes in the office. I'll go through them more
  thoroughly tomorrow morning.}\bcp{What the comment means is that names are
not usually formatted as ``Solomon, Maina'' but rather as ``Maina, Solomon''.}
%
\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

The denotation $\llbracket q \rrbracket$ of a QRE lens $q : Q_1 \Leftrightarrow
Q_2$ is a quotient lens $\llbracket q \rrbracket : W(Q_1)/{\eqrel{Q_1}}
\Longleftrightarrow W(Q_2)/{\eqrel{Q_2}}$. The typing rules and denotation of
QRE lenses are given in Figure~\ref{fig:qlenssemantics}.
\begin{figure}[ht]
\centering
\[
\begin{array}[b]{l@{\qquad}l}
\RuleSide{\ell : R \Leftrightarrow S}{\kw{lift}(\ell): id(R) \Leftrightarrow
id(S)}{\mathit{(Lift)}} &
\begin{array}[b]{r@{\ }c@{\ }l}
\llbracket \kw{lift}(\ell) \rrbracket.\get &=&  \llbracket \ell \rrbracket\\
\llbracket \kw{lift}(\ell) \rrbracket.\lput &=& \llbracket \ell \rrbracket^{-1}
\end{array}
\end{array}
\]

\[
\begin{array}[b]{c@{\qquad}c}
\RuleSide{q : Q_2  \Leftrightarrow Q_3 &
\wf{Q_1} &
K(Q_1) = W(Q_2)}
{\lquot(Q_1, q): Q_1 \; ; \; Q_2 \Leftrightarrow Q_3}{\mathit(Lquot)} &
\begin{array}[b]{r@{\ }c@{\ }l}
\llbracket \lquot(Q_1, q) \rrbracket.\get  &=& \llbracket q
\rrbracket.\get \circ \canonize(Q_1)\\
\llbracket \lquot(Q_1, q) \rrbracket.\lput &=& \llbracket q
\rrbracket.\lput
\end{array}
\end{array}
\]

\[
\begin{array}[b]{c@{\qquad}c}
\RuleSide{q : Q_1 \Leftrightarrow Q_3 & \wf{Q_3} & W(Q_3) = K(Q_2)}
{\rquot(q, Q_1):Q_1 \Leftrightarrow Q_2 \; ; \; Q_3}{\mathit{(Rquot)}} &
\begin{array}[b]{r@{\ }c@{\ }l}
\llbracket \rquot(q, Q_1) \rrbracket.\get  &=& \llbracket q
\rrbracket.\get\\
\llbracket \rquot(q, Q_1) \rrbracket.\lput &=& \llbracket q
\rrbracket.\lput \circ \canonize(Q_3)
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{q : Q_1 \Leftrightarrow Q_2 &
W(Q_1)^{*!},W(Q_2)^{*!} & K(Q_1)^{*!}, K(Q_2)^{*!}}
{{q}^* : {Q_1}^* \Leftrightarrow {Q_2}^*}{\mathit{(Star)}} &
\begin{array}[b]{r@{\ }c@{\ }l}
\llbracket {q}^* \rrbracket.\get  &=& (\llbracket q \rrbracket.\get)^*\\
\llbracket {q}^* \rrbracket.\lput &=& (\llbracket q \rrbracket.\get)^*
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{\substack{q_1 : Q_1 \Leftrightarrow Q_3 \\ q_2 : Q_2 \Leftrightarrow
Q_4} & {\substack{W(Q_1) \cdot^! W(Q_2)\\ K(Q_1) \cdot^! K(Q_2)}}
& {\substack{W(Q_3) \cdot^! W(Q_4)\\ K(Q_3) \cdot^! K(Q_4)}}}
{q_1 \cdot q_2: Q_1 \cdot Q_2 \Leftrightarrow Q_3 \cdot Q_4}{\mathit{(Concat)}}
&
\begin{array}[b]{r@{\ }c@{\ }l}
\llbracket q_1 \cdot q_2 \rrbracket.\get  &=& \llbracket q_1 \rrbracket.\get
\cdot \llbracket q_2 \rrbracket.\get\\
\llbracket q_1 \cdot q_2\rrbracket.\lput &=& \llbracket q_1 \rrbracket.\lput
\cdot \llbracket q_2 \rrbracket.\lput
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{\substack{q_1 : Q_1 \Leftrightarrow Q_3 \\ q_2 : Q_2 \Leftrightarrow
Q_4} & {\substack{W(Q_1) \cdot^! W(Q_2)\\ K(Q_1) \cdot^! K(Q_2)}}
& {\substack{W(Q_4) \cdot^! W(Q_3)\\ K(Q_4) \cdot^! K(Q_3)}}}
{q= q_1 \cdot q_2: Q_1 \cdot Q_2 \Leftrightarrow Q_4 \cdot
Q_3}{\mathit{(Swap)}} &
\begin{array}[b]{r@{\ }c@{\ }l}
\llbracket q \rrbracket.\get(s_1 \cdot s_2)  &=& \llbracket q_2
\rrbracket.\get(s_2) \cdot \llbracket q_1 \rrbracket.\get(s_1)\\
\llbracket q \rrbracket.\lput(t_2 \cdot t_1) &=&
\llbracket q_1 \rrbracket.\lput(t_1) \cdot \llbracket q_2 \rrbracket.\lput(t_2)
\end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\RuleSide{\substack{q_1 : Q_1 \Leftrightarrow Q_3 \\ q_2 : Q_2 \Leftrightarrow
Q_4} & \substack{\mathcal{L}(W(Q_1)) \cap \mathcal{L}(W(Q_2)) = \varnothing
\\ \mathcal{L}(W(Q_3)) \cap \mathcal{L}(W(Q_4)) = \varnothing }}
{q_1 \sep q_2: (Q_1 \sep Q_2)\Leftrightarrow (Q_3 \sep Q_4)}{\mathit{(Or)}} &
\begin{array}[b]{r@{\ }c@{\ }l}
\llbracket q_1 \sep q_2 \rrbracket.\get(s) =
\begin{cases}
\llbracket q_1 \rrbracket.\get (s) & \text{if } s \in \mathcal{L}(W(Q_1))\\
\llbracket q_2 \rrbracket.\get (s) & \text{if } s \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\llbracket q_1 \sep q_2 \rrbracket.\lput(s) =
\begin{cases}
\llbracket q_1 \rrbracket.\lput (s) & \text{if } s \in \mathcal{L}(W(Q_3))\\
\llbracket q_2 \rrbracket.\lput (s) & \text{if } s \in \mathcal{L}(W(Q_4))\\
\end{cases}
\end{array}
\end{array}
\]

@@

By the induction hypothesis, there exist bijective lenses,
$\ell_1 : K(Q_1) \Leftrightarrow K(Q_2)$ and $\ell_2 : K(Q_3) \Leftrightarrow
K(Q_4)$ such that:
\begin{center}
\begin{tabular}{l l}
$\llbracket q_1 \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(Q_1)$ & $\llbracket q_2 \rrbracket.\get = \llbracket \ell_2
\rrbracket \circ \canonize(Q_3)$ \\
$\llbracket q_1 \rrbracket.\lput = {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_2)$ & $\llbracket q_2 \rrbracket.\lput = {\llbracket \ell_2
\rrbracket}^{-1} \circ \canonize(Q_4)$
\end{tabular}
\end{center}
%
Consequently:
\begin{align*}
\llbracket q_2 \rrbracket.\get \circ \llbracket q_1 \rrbracket.\get &=
(\llbracket \ell_2 \rrbracket \circ \canonize(Q_3)) \circ (\llbracket \ell_1
\rrbracket \circ \canonize(Q_1))\\
&= \llbracket \ell_2 \rrbracket \circ (\canonize(Q_3) \circ \llbracket \ell_1
\rrbracket) \circ \canonize(Q_1)\\
&= (\llbracket \ell_2 \rrbracket \circ \llbracket \ell_1 \rrbracket) \circ
\canonize(Q_1)\\
&= \llbracket \ell_1 \; ; \; \ell_2 \rrbracket \circ \canonize(Q_1)
\end{align*}
We are permitted to claim the third step from the second since
$\canonize(Q_3)$ is the identity function on $K(Q_2)$ which is
syntactically equal to $K(Q_3)$ by assumption. A similar argument shows that:
$$\llbracket q_1 \rrbracket.\lput \circ \llbracket q_2 \rrbracket.\lput =
\llbracket \ell_1 \; ; \; \ell_2 \rrbracket^{-1} \circ
\canonize(Q_4)$$

@@

\section{Conclusion and Future Work}
\label{concl}

In this paper, we showed how to synthesize a class of bijective quotient lenses
using the bijective lens synthesis system Optician as a plug-in for our
synthesis algorithm. In order to achieve this, we first introduced {\em
Quotient regular languages} to specify regular languages with an equivalence
relation defined on them. Then, we introduced {\em QRE lenses}, which
are bijective quotient lenses that map between QREs via bijective lenses. We
proved a normal form theorem for QREs that enabled us to (1) extend the
synthesis algorithm used by Optician to synthesize QRE lenses, and (2) prove
that if there is a QRE lens that satisfies the input specification, then the
extended algorithm returns such a lens. Finally we tested QRE-enhanced optician
on the Optician benchmark suite and demonstrated that QREs can save programmers
significant effort compared to both Optician and basic Boomerang, and that
QRE-enhanced Optician maintains competitive performance over both of these
systems.

\BCP{Wrote the rest---comments?}
Looking ahead, we would like to investigate whether the techniques proposed
here can be applied to synthesizing quotiented variants of richer classes of
lenses, such as ``classic'' asymmetric lenses~\cite{FosterLenses} or
symmetric lenses~\cite{symmetriclenses}.  The main technical issues are
developing synthesis procedures for the unquotientied variants of a given
class---we are currently attacking the case of symmetric lenses---and
showing that quotiented lenses in this class can be normalized along the
lines of Theorem~\ref{normal form}.

Along a different dimension, we are experimenting with larger examples of
quotiented bijective lenses.  So far, the three ``specialized'' QRE
combinators (\kw{squash}, \kw{perm}, and \kw{collapse}) have proved
sufficient, but we expect eventually to encounter examples that will lead us
to consider adding further combinators to supplement these.
